---
title: County Revenues
author: Brian Bowling
date: '2019-03-05'
slug: county-revenues
categories:
  - counties
tags:
  - finances
---


```{r include=F}

library(tidyverse)
library(DT)

```

```{r include=F}


load('C:/Users/Brian/Documents/Projects/pa_local_gov/data_county_web.RData')

```

Here's a quick look at county revenues. It includes the percent that comes from local taxes as well as the percent that comes from intergovernmental transfers from federal, state and other local government agencies.



```{r echo=F}

data_county_web %>% 
  datatable(extensions= "FixedColumns",
            options=list(scrollX = TRUE, fixedColumns=list(leftColumns=2)),
            colnames=c('County','Total general fund revenues','From taxes','From federal','From state','From other local')) %>% 
  formatCurrency(3,digits=0) %>%
  formatPercentage(4:7,digits=1)

```

The data comes from the annual reports the counties file with the state Department of Community and Economic Development and can be found, in more detail, here: http://munstats.pa.gov/Reports/ReportInformation2.aspx?report=cAfrForm

I plan to post some of the spending data next as well as a look at how each county's debt compares to its total revenues and tax revenues. Down the line, I plan to do the same with municipalities but their reports follow a different format that I have yet to decipher.
