---
title: West Virginia has a higher density of churches than Pennsylvania
author: Brian Bowling
date: '2019-12-28'
slug: west-virginia-has-more-churches-per-10-square-miles-than-pennsylvania
categories:
  - misc
tags:
  - geo
---

```{r prep, include=F}

library(tidyverse)
library(DT)
library(usmap)
library(viridis)

path_data_census <- 'C:/Users/bowli/Documents/data/Census/geo_files/'
path_common <- 'C:/Users/bowli/Documents/data/commonly_used/'

national_names <- read_delim(paste0(path_data_census,'NationalFile/NationalFile_20191101.txt'),
                             delim='|',
                             col_types = cols(.default=col_character(),
                                  ELEV_IN_M = col_double(),
                                  ELEV_IN_FT = col_double()),
                             quote = "")


# state names
state_names <- read_tsv(paste0(path_common,'state_abbreviations.csv'))

# state areas
state_areas <- read_csv(paste0(path_common,'state_areas.csv'),
                        col_types = cols(.default=col_double(),
                                         geography = col_character(),
                                         internal_point_latitude_value = col_character(),
                                         internal_point_longitude_value= col_character()))


state_areas <- state_areas %>% 
  mutate(geography = ifelse(geography=='U.S. Virgin Islands','Virgin Islands',geography))

state_areas <- left_join(state_names, state_areas, by=c('state'='geography'))


class_count <- national_names %>% 
  group_by(FEATURE_CLASS) %>% 
  summarise(Number = n()) %>% 
  arrange(desc(Number)) %>% 
  select(Class = FEATURE_CLASS, Number)

church_names <- national_names %>% 
  filter(FEATURE_CLASS == 'Church') %>% 
  group_by(FEATURE_NAME) %>% 
  summarise(Number = n()) %>% 
  top_n(50,wt=Number) %>% 
  arrange(desc(Number)) %>% 
  select(Name = FEATURE_NAME, Number)


church_number = nrow(national_names %>%
                       filter(FEATURE_CLASS == 'Church'))

church_names_count <- nrow(national_names %>%
                              filter(FEATURE_CLASS == 'Church') %>%
                              distinct(FEATURE_NAME))


church_names_unique <- nrow(national_names %>% 
                              filter(FEATURE_CLASS == 'Church') %>% 
                              group_by(FEATURE_NAME) %>% 
                              count() %>% 
                              filter(n==1))



churches_per_state <- national_names %>% 
  filter(FEATURE_CLASS=='Church') %>% 
  group_by(STATE_ALPHA) %>% 
  summarise(Number = n())

churches_per_state <- left_join(state_areas,churches_per_state, by=c('abbreviation'='STATE_ALPHA')) %>% 
  select(state, total_sq_mi, Number) %>% 
  mutate(churches_10sq_mi = round(Number / (total_sq_mi/10), digits=3))


pa_density <- round(churches_per_state$churches_10sq_mi[churches_per_state$state=="Pennsylvania"],digits=1)
wv_density <- round(churches_per_state$churches_10sq_mi[churches_per_state$state=="West Virginia"],digits=1)


```

Although Pennsylvania has nearly twice the land area of West Virginia, their number of churches is almost the same, according to Census and U.S. Geological Survey data.

Oddly enough, the land area values come from the Census while the number of churches comes from the Geological Survey's domestic names database, https://www.usgs.gov/core-science-systems/ngp/board-on-geographic-names/domestic-names.

I started taking a look at the database Friday (Dec. 27, 2019) and decided to see if the data confirmed my personal experience that West Virginia seemed to have more churches per square mile than Pennsylvania or pretty much any other place I'd ever lived.

After the initial look, I decided to look at churches per 10 square miles to make the numbers a little easier to grasp. The analysis confirmed my main suspicion (`r wv_density` churches for every 10 square miles in West Virginia versus `r pa_density` in Pennsylvania). I was surprised to find that it doesn't have higher concentrations than two other places I've lived: Connecticut and Virginia.


```{r echo=F}

unmappable = c('guam','puerto rico','virgin islands')

churches_per_state <- churches_per_state %>%
  mutate(category = cut(
    churches_10sq_mi,
    breaks=c(0,.5, 1, 2, 3, 102),
    labels = c('.5 or less', '.51 to 1', '1.1 to 2', '2.1 to 3', 'more than 3'),
    include.lowest = T
  ))


plot_usmap(regions='states',data=churches_per_state,values='category') +
  labs(title = "Church Concentrations by state") +
  theme(legend.position = 'right',
        plot.title=element_text(hjust=.5, size=24)) +
  scale_fill_viridis(name='Churches per 10 square miles',discrete=T,direction=-1)


```


One caveat, a notice on the U.S.G.S. site says that it quit adding the names of man-made structures to the database in 2014, so it doesn't cover any churches added in the last six years.

To give the issue some perspective, there are more churches in the database than any other natural or man-made feature. The `r prettyNum(church_number,big.mark=",")` churches make up about 10 percent of the entire database of about 2.3 million named features.


The table below gives the number of churches and the number per 10 square miles for all the states. 

```{r echo=F}

churches_per_state %>% 
  mutate(churches_10sq_mi = round(churches_10sq_mi,digits=1)) %>% 
  select(State = state, Churches=Number, Density = churches_10sq_mi) %>% 
  datatable() %>% 
  formatCurrency('Churches',currency='',digits=0)

```

