---
title: "Popularity of dog names in Allegheny County"
author: "Brian Bowling"
date: '2019-02-19'
slug: popularity-of-dog-names-in-allegheny-county
tags: pets
categories: misc
---

```{r prepwork, include=FALSE, warning=F,message=F}

library(tidyverse)
library(kableExtra)
library(lubridate)
library(formattable)
library(stringr)

load('C:/Users/Brian/Documents/Projects/dognames/dog_licenses_annual.RData')

# In some cases, no name was given and the county isn't consistent about how it
# codes this
dog_years <- dog_years %>% 
  mutate(DogName = ifelse(str_detect(DogName," NAME"),NA,DogName)) %>% 
  mutate(DogName = ifelse(str_detect(DogName,"NAME "),NA,DogName))



licenses_per_year <- dog_years %>% 
  group_by(year) %>% 
  summarise(licenses = n(),
            unique_names = n_distinct(DogName))

licenses_by_gender_year <- dog_years %>% 
  group_by(year,gender) %>% 
  summarise(licenses=n())

# when do people typically license their dogs?
license_month <- dog_years %>% 
  group_by(month = month(ValidDate)) %>% 
  summarise(licenses = n()) %>% 
  mutate(month_pct = licenses/sum(licenses))



# get the top 3 dog names by gender for each year
top_dogs <- dog_years %>% 
  filter(!is.na(gender)) %>% 
  group_by(year, gender, DogName) %>% 
  summarise(dogs=n()) %>% 
  top_n(3,dogs)

# get list of those names
top_male_names <- unique(top_dogs$DogName[which(top_dogs$gender=='Male')])

top_female_names <- unique(top_dogs$DogName[which(top_dogs$gender=='Female')])

top_all_names <- unique(top_dogs$DogName)

# create graphs of top male and female dogs by number of licenses per year
top_males_year <- dog_years %>% 
  filter(DogName %in% top_male_names) %>% 
  group_by(year, DogName) %>% 
  summarise(licenses=n()) %>% 
  ggplot(aes(x=year, y=licenses, group=DogName,color=DogName)) +
  geom_line() +
  scale_colour_brewer(palette='Set1') +
  scale_x_continuous(breaks=c(2007:2018)) +
  labs(title='Top male dog names',
       y='Number of licenses',
       x='Year',
       color='Name') +
  theme(text=element_text(family='Helvetica'),
        plot.title=element_text(size=24, hjust=0.5))


top_females_year <- dog_years %>% 
  filter(DogName %in% top_female_names) %>% 
  group_by(year, DogName) %>% 
  summarise(licenses=n()) %>% 
  ggplot(aes(x=year, y=licenses, group=DogName,color=DogName)) +
  geom_line() +
  scale_colour_brewer(palette='Set1') +
  scale_x_continuous(breaks=c(2007:2018)) +
  labs(title='Top female dog names',
       y='Number of licenses',
       x='Year',
       color='Name') +
  theme(text=element_text(family='Helvetica'),
        plot.title=element_text(size=24, hjust=0.5))


# create tables of male and female names by count with most frequent in 2017 as
# sort variable
top_males_table <- dog_years %>% 
  filter(DogName %in% top_male_names) %>% 
  group_by(year, DogName) %>% 
  summarise(licenses=n()) %>% 
  spread(year, licenses) %>% 
  arrange(desc(`2017`))

top_females_table <- dog_years %>% 
  filter(DogName %in% top_female_names) %>% 
  group_by(year, DogName) %>% 
  summarise(licenses=n()) %>% 
  spread(year, licenses) %>% 
  arrange(desc(`2017`))


```

Bella has usurped Molly but Buddy can't be beat when it comes to the most frequent names on dog licenses in Allegheny County.

```{r echo=FALSE,warning=F}

top_females_year

```


```{r echo=FALSE, warning=F}

top_males_year

```

The data comes from the Western Pennsylvania Regional Data Center, https://data.wprdc.org/dataset/allegheny-county-dog-licenses. It includes dogs licensed within the county but not those licensed inside the city of Pittsburgh.

County dog owners like a lot of variety in their pets' names. While top names for each gender accont for about 400 to 500 names, about 30,000 dogs are licensed every year with the number of unique names ranging from 7,031 in 2016 to 8,929 in 2008.

```{r echo=F}
licenses_per_year %>% 
  mutate(licenses = formattable::comma(licenses,digits=0),
         unique_names = formattable::comma(unique_names,digits=0)) %>% 
  kable(col.names=c("Year","Number of licenses","Number of unique names")) %>% 
  kable_styling(bootstrap_options=c('striped','responsive'),full_width = F)

```

The 2018 numbers are only for dogs licensed through August, but the data shows that most people license their dogs between December and May. Since the county's dog license site suggests they can send their dog license application in the same envelope as their property tax payment, which is due by April 30, this makes sense.

```{r echo=F}

license_month %>% 
  mutate(month=ifelse(month<13,month.name[month],"NA"),
         licenses=formattable::comma(licenses,digits=0),
         month_pct=formattable::percent(month_pct,digits=1)) %>% 
  kable(col.names=c("Month",'Licenses','Percent of total')) %>% 
  kable_styling(bootstrap_options=c('striped','responsive'),full_width = F)


```

