---
title: It's a nice place to live but I'm not sure I want to stay here
author: Brian Bowling
date: '2019-06-27'
slug: it-s-a-nice-place-to-live-but-i-m-not-sure-i-want-to-stay-here
categories:
  - misc
tags:
  - surveys
---

```{r libraries, include=F}

library(tidyverse)
library(haven)
library(DT)
library(scales)
library(janitor)

```


```{r, include=F}
local_news <- read_sav('c:/users/brian/documents/mystuff/projects/api_stuff/data_raw/Local News Public.sav', user_na = T) %>% 
  as_factor()

```


From the Pew Research Center's "2018 Local News Survey" is an interesting example of how the answer to a data analysis changes when you flip the question.


```{r quick_look function, include=F}

quick_look <- function(variable) {
  variable = toupper(variable) # make sure variable is uppercase
  local_news %>% 
    group_by(.dots = quo_name(variable)) %>% 
    summarise(weighted_n = sum(WEIGHT)) %>% 
    mutate(percent = weighted_n/sum(weighted_n)) %>% 
    select(-weighted_n) %>% 
    datatable(options=list(dom='t')) %>% 
    formatPercentage(2)
}

```


The data comes from the first two questions, which asked 34,897 people to rate the quality of living in their communities and how attached they feel to those communities.

LOCALRATING: Overall, how would you rate your community as a place to live?

```{r localrating, echo=F}

quick_look('LOCALRATING')

```

COMATTACH: In general, how attached do you feel to your local community?

```{r comattach, echo=F}

quick_look('COMATTACH')

```

As you might suspect, people who feel very attached to their communities tend to give them favorable ratings.

```{r rating by attachment, echo=F}

sketch = htmltools::withTags(table(
  class='display',
  thead(
    tr(
      th(rowspan=2,'Community Rating'),
      th(colspan=5,'How attached to the community')
    ),
    tr(
      th(colspan=1,'Very'),
      th(colspan=1,'Somewhat'),
      th(colspan=1,'Not very'),
      th(colspan=1,'Not at all'),
      th(colspan=1,'Refused')
    )
  )
))


local_news %>% 
  group_by(LOCALRATING, COMATTACH) %>% 
  summarise(weighted_n = sum(WEIGHT)) %>% 
  group_by(COMATTACH) %>% 
  mutate(group_wt = sum(weighted_n)) %>% 
  mutate(percent = weighted_n/group_wt) %>% 
  select(-weighted_n, -group_wt) %>% 
  spread(COMATTACH, percent) %>% 
  adorn_totals(where='row') %>% 
  datatable(container=sketch, rownames=F,options = list(dom = 't')) %>% 
  formatPercentage(2:6)


```

But what may suprise you is that people who give their communities good ratings may not necessarly feel that attached to them.

```{r attach by rating, echo=F}



sketch = htmltools::withTags(table(
  class='display',
  thead(
    tr(
      th(rowspan=2,'How attached'),
      th(colspan=5,'Quality of living in the community')
    ),
    tr(
      th(colspan=1,'Excellent'),
      th(colspan=1,'Good'),
      th(colspan=1,'Only fair'),
      th(colspan=1,'Poor'),
      th(colspan=1,'Refused')
    )
  )
))

local_news %>% 
  group_by(LOCALRATING, COMATTACH) %>% 
  summarise(weighted_n = sum(WEIGHT)) %>% 
  group_by(LOCALRATING) %>% 
  mutate(group_wt = sum(weighted_n)) %>% 
  mutate(percent = weighted_n/group_wt) %>% 
  select(-weighted_n, -group_wt) %>% 
  spread(LOCALRATING, percent) %>% 
  adorn_totals(where='row') %>% 
  datatable(container=sketch, rownames=F,options = list(dom = 't')) %>% 
  formatPercentage(2:6)


```
