---
title: Data reporting - average percent change
author: Brian Bowling
date: '2020-06-08'
slug: data-reporting-average-percent-change
categories:
  - data_reporting
tags:
  - stats
---

# Calculating the rate of change

When it comes to analyzing several observations over time, figuring out the average percent increase or decrease (i.e. growth rate) is a useful but potentially misleading tool. Looking over a city's budget for the past five years, for example, you can see how the growth in spending and revenues compared to each other as well as inflation. Comparing county population growth rates over a few decades can help pinpoint the ones with consistent growth.

Here are a few examples to illustrate the point.

```{r libraries, message=F}

library(tidyverse) # general data-wrangling package
library(DT) # helps create readable data tables

```

First, a look at Pittsburgh's budgets as provided by the city on its website:
https://pittsburghpa.gov/omb/budgets-reports

Note: This is the budgeted amount from each year, not the actual revenues and spending. For that you would want 

```{r pittsburgh data}

city_budget <- tribble(
  ~Year,~Revenue,~Expenditure,
2020, 608235885, 608100046,
2019, 582342954, 574901207,
2018, 566481637, 556414971,
2017, 545427808, 539193140,
2016, 525005696, 518983192,
2015, 516550159, 507817535
)


```

## Total percent change

The total percent change is easy. You just subtract the 2015 value from the 2020 value and then divide the result by the 2015 value.

```{r total percent change}

city_budget %>% 
  filter(Year %in% c(2020,2015)) %>% 
  pivot_longer(cols=c(Revenue, Expenditure),names_to="Category") %>% 
  pivot_wider(id_cols=Category, names_from=Year) %>% 
  mutate(`Percent change` = round((`2020`-`2015`)/`2015`,digits=3)) %>% 
  datatable(options=list(dom='t')) %>% 
  formatPercentage('Percent change',digits=1) %>% 
  formatCurrency(c('2020','2015'),digits=0)

```

## Growth rate

Is the 19.7% growth in spending a lot or a little? You could answer that by running the same calculations for other cities and even counties and states to see how Pittsburgh compares to them. A quicker answer, however, would come from figuring out the annual growth rate over the past five years since most people have an idea of how their finances have changed annually.

The growth rate calculation is more complicated than the percent change.

You divide the 2020 value by the 2015 value and then you raise that result by the inverse of the number of periods in between the two numbers and then subtract one from the result. In this case, that the number of periods is five years since we're looking for an annual growth rate. You might think you could change periods to 60 to get the monthly growth rate, but that would assume that the 2015 and 2020 values were monthly values.


```{r growth rate}


city_budget %>% 
  filter(Year %in% c(2020,2015)) %>% 
  pivot_longer(cols=c(Revenue, Expenditure),names_to="Category") %>% 
  pivot_wider(id_cols=Category, names_from=Year) %>% 
  mutate(`Growth rate` = round((`2020`/`2015`)**(1/(2020-2015))-1,digits=3)) %>% 
  datatable(options=list(dom='t')) %>% 
  formatPercentage('Growth rate',digits=1) %>% 
  formatCurrency(c('2020','2015'),digits=0)


```

Since the annual inflation rate has been hovering around 2% for that same period, you can see that Pittsburgh's growth in revenue and spending is a little above average. That's not necessarily bad since it could indicate the city's economy is doing better than average, but pending growing faster than the revenue is a concern.

## Cherry picking

Cherry picking, in this context, consists of picking your starting point to get the growth rate you want to make your point.

The overall budget totals are a little too regular to illustrate this, so I'm going to drill down to spending on "General Government" activity from 2000 to 2019, which is found in the city's 2019 Comprehensive Annual Financial Report.
https://pittsburghpa.gov/controller/cafr

Here's a line chart of the numbers:

```{r}

gen_gvt_spending <- tribble(
  ~Year, ~General_government,
2019, 104814151,
2018, 107097053,
2017, 79890619,
2016, 90146429,
2015, 92461452,
2014, 74786585,
2013, 84139729,
2012, 76848176,
2011, 72151719,
2010, 71614406
)

gen_gvt_spending %>% 
  ggplot(aes(x=Year, y=General_government)) +
  geom_line() +
  scale_y_continuous(labels=scales::dollar, limit=c(0,NA)) +
  scale_x_continuous(breaks=c(2010:2019)) +
  labs(title="Pittsburgh spending on general government activity",
       caption="Source: Comprehensive Annual Financial Report",
       y="",
       x="")
```

Here's what the annual growth rate looks like in 2019 depending on which year you use as the base.

```{r}

ref_amt = gen_gvt_spending$General_government[gen_gvt_spending$Year==2019]

gen_gvt_spending %>% 
  mutate(growth_rate = round((ref_amt/General_government)**(1/(2019-Year))-1,digits=3)) %>% 
  select('Base Year'=Year, `2019 Growth rate` = growth_rate) %>% 
  datatable(options=list(dom='t'),rownames = F, width=200) %>% 
  formatPercentage('2019 Growth rate',digits=1)

```
If you wanted to make the growth rate look outrageous, you would probably cherry pick 2017 as your base. Of course, if you were including this in the same analysis at the five-year look at budget totals, the 2017 reference year would look suspicious.

If you're trying to be honest with your analysis, the different ending years present a problem. Do you use 2019 for both or use 2020 for the budget totals and 2019 for the General Government activity spending with a note explaining the 2020 number isn't available yet? Do you back up the General Government spending analysis to 2014 so that you have five-year trends for both?

There really is no right answer to those questions. It's a matter of preference as long as your transparent about your choices and the reasons behind them.

