---
title: Monty Hall Problem
author: Brian Bowling
date: '2020-11-09'
slug: monty-hall-problem
categories:
  - misc
tags:
  - stats
---

Some people intuitively grasp the correct answer to the Monty Hall Problem. Others, myself included, reluctantly accept the answer without really understanding it.

Last night, for some reason, I had a breakthrough.

For those of you not familiar with it, the problem is named after the former host of Let's Make A Deal, Monty Hall, and deals with a contestant having to choose one of three doors.

Behind one of the doors is a new car. Behind each of the other two doors is a goat. If the contestant picks the door with the car, he or she wins the car.

After the contestant picks a door, Monty Hall opens one of the other doors to show that it has a goat behind it. The contestant then has the choice of staying with the door he or she has already picked or switching to the remaining door.

While it seems like there's a 50-50 chance of the car being behind either door, that's not correct. In fact, the odds of the car being behind the door the contestant originally picked is only 1 in 3 while the odds for it being behind the other door are 2 in 3, so the best strategy here is for the contestant to switch to the other door.

Not convinced? Let's run some simulations showing what happens when two different contestants play the game 1,000 times each. One of the players (the switcher), always chooses to switch to the remaining door while the other (the stayer) sticks with his or her original selection.


```{r monty, message=FALSE}

# load libraries
library(tidyverse)

# constants
num_rounds <- 1000

# function to simulate playing the game

monty_runs <- function(switcher, num_rounds) {
  # table to hold results of games
  player_results = tribble(
    ~result, ~count,
    'goat', 0,
    'car',0
  )

  # start of game loop
  for (i in 1:num_rounds){
  
    # pick the door for the car
    car <- sample(1:3,1)
  
    # have player pick a door
    player_choice = sample(1:3,1)
    
    # make a list of the doors player hasn't picked
    other_doors <- 1:3
    other_doors <- other_doors[other_doors != player_choice]
    
    # pick a door to open, revealing it has a goat behind it

    # if player picked the door that has a car behind it, randomly open
    # one of the other doors
    # if player picked a door with a goat behind it, open the other door with
    # a goat behind it
    if (player_choice==car) {
      reveal = other_doors[sample(1:2,1)]
    } else {
      reveal = ifelse(other_doors[1]==car,other_doors[2],other_doors[1])
    }
    
    # make a list of the doors that are still closed
    doors_left <- 1:3
    doors_left <- doors_left[doors_left != reveal]
    
    # have player decided whether to stick with the door he or she originally
    # picked or switch to the other closed door
    if (switcher) {
      player_choice <- doors_left[doors_left != player_choice]
    }
    
    # update results
    player_wins <- ifelse(player_choice==car,'car','goat')
    
    player_results <- player_results %>% 
        mutate(count = ifelse(result==player_wins,count+1, count))
  }
  return(player_results)
}

# run the simulation twice, once for a player who always switches
# and once for a player who always sticks and compile the results

player_switches <- monty_runs(TRUE,1000) %>% 
  mutate(player_type='switcher')

player_stays <- monty_runs(FALSE, 1000) %>% 
  mutate(player_type='stayer')

monty_results <- bind_rows(player_switches, player_stays) %>% 
  pivot_wider(id_cols=player_type,names_from=result,values_from=count) %>% 
  mutate(pct_car = round(car/(goat+car),digits=3))

```

Here are the results for the two types of player.

```{r results, message=FALSE}

library(DT)

monty_results %>% 
  datatable(colnames = c("Player Style","Picks a goat","Picks the car","Success percentage"),
            options=list(dom='t')) %>% 
  formatPercentage('pct_car')

```

While the percentages vary each time through the simulation, you can see that the switcher gets the car about twice as often as the stayer.

Still not convinced, then let me try to explain it.

When the contestant makes the original choice, he or she has a 2 in 3 chance of picking a door with a goat because two of the three doors have goats behind them.

Monty Hall knows what's behind each door, so his selection of another door that has a goat behind it isn't random and doesn't affect the odds that the contestant picked a door with a goat behind it. If the contestant picked the door with a car behind it, Monty simply picks one of the other doors and opens it. If the contestant picked a door with a goat behind it, Monty's only choice is to open the other door with a goat behind it.

So two out of three times, the contestant picks a door with a goat behind it and Monty opens the other door with a goat behind it. That means two out of three times, the door Monty didn't open is the one with the car behind it.

Seems simple to me now, but I've spent a lot of time and read a lot of posts in various forums arguing against this correct answer.