---
title: County-by-county Covid-19 trends
author: Brian Bowling
date: '2020-05-09'
slug: local-covid-19-trends
categories:
  - misc
tags:
  - stats
---

```{r prep, include=F}

# libraries

library(tidyverse)
library(janitor)
library(lubridate)
library(patchwork)

# parameters

swpa <- c("Allegheny County","Armstrong County","Beaver County","Butler County",
          "Fayette County","Greene County","Lawrence County","Washington County",
          "Westmoreland County")


time_zone <- "America/New_York"
fmt <- "%m/%d/%Y"
start_date <- as_date("4/1/2020",tz=time_zone, format=fmt)

# load data

path_raw_data <- 'D:/Documents/projects/data_workshop/data_raw/covid/covid_casesdeaths_reshaped_bln.csv'

local_data <- read_csv(path_raw_data)

local_data <- local_data %>% 
  clean_names()

# convert date string to date variable

local_data <- local_data %>% 
  mutate(date = as.Date(date, '%m/%d/%y'))

# create tables for latest totals

latest <- local_data %>% 
  filter(date == max(date))

pa_latest <- latest %>% 
  filter(state=="PA")

swpa_latest <- pa_latest %>% 
  filter(county_name %in% swpa)

latest_table <- bind_rows(
  latest %>% 
    group_by(date) %>% 
    summarise(cases = sum(cases),
              deaths = sum(deaths)) %>% 
    mutate(name = "US total"),
  pa_latest %>% 
    group_by(date) %>% 
    summarise(cases = sum(cases),
              deaths = sum(deaths)) %>% 
    mutate(name = "PA total"),
  swpa_latest %>% 
    group_by(date) %>% 
    summarise(cases=sum(cases),
              deaths=sum(deaths)) %>% 
    mutate(name="SWPA total"),
  swpa_latest %>% 
    rename(name=county_name)) %>% 
  select(name, date, cases, deaths)

# calculate daily numbers

local_data <- local_data %>% 
  group_by(state, county_name) %>% 
  mutate(new_cases = cases - lag(cases, order_by = date),
         new_deaths = deaths - lag(deaths, order_by = date))

# subset data for easier graphing

us_sum <- local_data %>% 
  filter(date >= start_date) %>% 
  group_by(date) %>% 
  summarise(new_cases = sum(new_cases),
            new_deaths = sum(new_deaths))

pa_sum <- local_data %>% 
  filter(date >= start_date, state=="PA") %>% 
  group_by(date) %>% 
  summarise(new_cases = sum(new_cases),
            new_deaths = sum(new_deaths))

swpa_data <- local_data %>% 
  filter(date >= start_date, state=="PA", county_name %in% swpa)

swpa_sum <- swpa_data %>% 
  group_by(date) %>% 
  summarise(new_cases = sum(new_cases),
            new_deaths = sum(new_deaths))

```

# Finding local trend data

While it's easy to find county-level totals for the number of Covid-19 cases and deaths, it's not as easy to find trend data. Fortunately, the nonproft USAFacts.org and Big Local News, a program in Stanford University’s Journalism and Democracy Initiative, have collaborated to provide that kind of data. Here's the latest data for nine counties in Southwestern Pennsylvania: Allegheny, Armstrong, Beaver, Butler, Fayette, Greene, Lawrence, Washington and Westmoreland.


```{r swpa, echo=F}

case_num <- latest_table %>% 
  filter(name=="SWPA total") %>% 
  pull(cases)

death_num <- latest_table %>% 
  filter(name=="SWPA total") %>% 
  pull(deaths)

subtitle_text <- paste0("Total cases: ", prettyNum(case_num,big.mark = ','),
                        "  Total deaths: ", prettyNum(death_num, big.mark=","))


swpa_cases <- swpa_sum %>% 
  ggplot(aes(x=date, y=new_cases)) +
  geom_col(fill='yellow',color='black') +
  labs(title="New cases",
       x="",
       y="") +
  scale_y_continuous(label=scales::label_comma(accuracy=1))

swpa_deaths <- swpa_sum %>% 
  ggplot(aes(x=date, y=new_deaths)) +
  geom_col(fill='red',color='black') +
  labs(title="New deaths",
       x="",
       y="") +
  scale_y_continuous(label=scales::label_comma(accuracy=1))

(swpa_cases | swpa_deaths) + plot_annotation(title="Covid-19 trends for Southwestern PA",
                                             subtitle=subtitle_text,
                                         caption="Includes Allegheny, Armstrong, Beaver, Butler, Fayette, Greene, Lawrence, Washington and Westmoreland counties.\nSource:USAFacts.org",
                                         theme = theme(plot.title = element_text(size = rel(2),
                                                                                 hjust=0.5),
                                                       plot.subtitle=element_text(hjust=0.5)))


```

At the bottom of this post are the graphs for each individual county but, first, here are the bigger pictures.


```{r us, echo=F}

case_num <- latest_table %>% 
  filter(name=="US total") %>% 
  pull(cases)

death_num <- latest_table %>% 
  filter(name=="US total") %>% 
  pull(deaths)

subtitle_text <- paste0("Total cases: ", prettyNum(case_num,big.mark = ','),
                        "  Total deaths: ", prettyNum(death_num, big.mark=","))

us_cases <- us_sum %>% 
  ggplot(aes(x=date, y=new_cases)) +
  geom_col(fill='yellow',color='black') +
  labs(title="New cases",
       x="",
       y="") +
  scale_y_continuous(label=scales::label_comma(accuracy=1))

us_deaths <- us_sum %>% 
  ggplot(aes(x=date, y=new_deaths)) +
  geom_col(fill='red',color='black') +
  labs(title="New deaths",
       x="",
       y="") +
  scale_y_continuous(label=scales::label_comma(accuracy=1))

(us_cases | us_deaths) + plot_annotation(title="Covid-19 trends for the United States",
                                         subtitle=subtitle_text,
                                         caption="Source: USAFacts.org",
                                         theme = theme(plot.title = element_text(size = rel(2),
                                                                                 hjust=0.5),
                                                       plot.subtitle=element_text(hjust=0.5)))


```


```{r pa, echo=F}


case_num <- latest_table %>% 
  filter(name=="PA total") %>% 
  pull(cases)

death_num <- latest_table %>% 
  filter(name=="PA total") %>% 
  pull(deaths)

subtitle_text <- paste0("Total cases: ", prettyNum(case_num,big.mark = ','),
                        "  Total deaths: ", prettyNum(death_num, big.mark=","))


pa_cases <- pa_sum %>% 
  ggplot(aes(x=date, y=new_cases)) +
  geom_col(fill='yellow',color='black') +
  labs(title="New cases",
       x="",
       y="") +
  scale_y_continuous(label=scales::label_comma(accuracy=1))

pa_deaths <- pa_sum %>% 
  ggplot(aes(x=date, y=new_deaths)) +
  geom_col(fill='red',color='black') +
  labs(title="New deaths",
       x="",
       y="") +
  scale_y_continuous(label=scales::label_comma(accuracy=1))

(pa_cases | pa_deaths) + plot_annotation(title="Covid-19 trends for Pennsylvania",
                                         subtitle=subtitle_text,
                                         caption="Source: USAFacts.org",
                                         theme = theme(plot.title = element_text(size = rel(2),
                                                                                 hjust=0.5),
                                                       plot.subtitle=element_text(hjust=0.5)))

```


```{r county_graph, include=F}


graph_county <- function(county, data) {
  title_text <- paste("Covid-19 trends for",county)
  
  case_num <- data %>% 
    filter(name==county) %>% 
    pull(cases)
  
  death_num <- latest_table %>% 
    filter(name==county) %>% 
    pull(deaths)

  subtitle_text <- paste0("Total cases: ", prettyNum(case_num,big.mark = ','),
                          "  Total deaths: ", prettyNum(death_num, big.mark=","))

  county_cases <- swpa_data %>%
    filter(county_name==county) %>% 
    ggplot(aes(x=date, y=new_cases)) +
    geom_col(fill='yellow',color='black') +
    labs(title="New cases",
         x="",
         y="") +
    scale_y_continuous(label=scales::label_comma(accuracy=1))
  
  county_deaths <- swpa_data %>%
    filter(county_name==county) %>% 
    ggplot(aes(x=date, y=new_deaths)) +
    geom_col(fill='red',color='black') +
    labs(title="New deaths",
         x="",
         y="") +
    scale_y_continuous(label=scales::label_comma(accuracy=1))
  
  
  (county_cases | county_deaths) + plot_annotation(title=title_text,
                                                   subtitle=subtitle_text,
                                                   caption="Source: USAFacts.org",
                                                   theme = theme(plot.title = element_text(size = rel(2),
                                                                                           hjust=0.5),
                                                                 plot.subtitle=element_text(hjust=0.5)))
}

```



```{r allegheny, echo=F}

graph_county("Allegheny County",latest_table)

```

```{r armstrong, echo=F}

graph_county("Armstrong County", latest_table)

```

```{r beaver, echo=F}

graph_county("Beaver County",latest_table)

```

```{r butler, echo=F}

graph_county("Butler County",latest_table)


```

```{r fayette, echo=F}

graph_county("Fayette County",latest_table)


```

```{r greene, echo=F}

graph_county("Greene County",latest_table)

```

```{r lawrence, echo=F}

graph_county("Lawrence County",latest_table)

```

```{r washington, echo=F}

graph_county("Washington County",latest_table)

```

```{r westmoreland, echo=F}

graph_county("Westmoreland County",latest_table)

```

