---
title: U.S above median for deaths from coronavirus infections
author: Brian Bowling
date: '2020-04-15'
slug: u-s-above-median-for-deaths-from-coronavirus-infections
categories:
  - misc
tags:
  - stats
---


```{r generate stats, include=F}

# libraries
library(tidyverse)
library(lubridate)
library(DT)

# get the data
data_raw_path <- 'D:/Documents/projects/data_workshop/data_raw/'

# get last download date-time
load(paste0(data_raw_path,'dl.RData'))

ecdpc_url <- "https://opendata.ecdc.europa.eu/covid19/casedistribution/csv"
current_dt <- now('UTC')

if (as.numeric(current_dt - dl_dt,unit='days') > 1) {
  # been more than 24 hours since last download, so get new data
  download.file(ecdpc_url,paste0(data_raw_path,'casedistribution.csv'))
  # save current as new download date-time
  dl_dt <- current_dt
  save(dl_dt,file=paste0(data_raw_path,'dl.RData'))
}

ecdpc_data <- read_csv(paste0(data_raw_path,'casedistribution.csv'))

# convert dateRep to date

ecdpc_data <- ecdpc_data %>% 
  mutate(dateRep=as.Date(dateRep,'%d/%m/%Y'))

survival <- ecdpc_data %>% 
  group_by(countriesAndTerritories) %>% 
  summarise(cases_total = sum(cases),
      deaths_total = sum(deaths)) %>%
  filter(cases_total >= 100) %>% 
  mutate(death_rate = round(deaths_total/cases_total,digits=3))

survival_quartiles <- survival %>% 
  filter(death_rate > 0) %>% 
  summarise(best = min(death_rate),
            q1 = quantile(death_rate,.25),
            median = median(death_rate),
            q3 = quantile(death_rate,.75),
            worst = max(death_rate),
 )

us_stats <- survival %>% 
  filter(countriesAndTerritories == 'United_States_of_America')

country_count <- survival %>% 
  filter(deaths_total > 0) %>% 
  count() %>% 
  pull(n)

# graph it
p1 <- survival %>% 
  filter(death_rate > 0) %>%
  arrange(death_rate) %>% 
  ggplot(aes(x=death_rate)) +
  annotate('text', x=.045, y=26, label='US (4.3%)',color='red') +
  annotate('text',x=.028, y=26, label='Median', color='blue') +
  geom_histogram(breaks = seq(0,.16,.01),fill='yellow',color='black') +
  geom_segment(aes(x=us_stats$death_rate,y=0,xend=us_stats$death_rate,yend=25),color='red') +
  geom_segment(aes(x=.028, y=0, xend=.028, yend=25),color='blue') +
  labs(title='Deaths per cases',
       subtitle="(Includes 125 countries reporting at least one death from Covid-19)",
       y='Number of countries',
       x= 'Deaths as percentage of cases',
       caption = "Source: European Centre for Disease Prevention and Control") +
  scale_x_continuous(labels=scales::percent,
                     breaks = c(0.002,.014, .028,.048, .157)) +
  theme(plot.title=element_text(hjust=0.5,size=rel(1.5)),
        plot.subtitle = element_text(hjust=0.5))

```

The United States is seeing a higher rate of deaths from coronavirus infections than the median for the 125 countries that have reported cases and deaths to the European Centre for Disease Prevention and Control, according to today's stats.

As with most of coronavirus numbers, the main caveat is that it's hard to gauge the honesty and competency of any of the countries, including the United States, when it comes to detecting infections and deaths related to Covid-19. It's relatively safe to assume that the actual number of cases and deaths is higher for most, if not all the countries.

But, for what it's worth, here's the breakdown.

```{r echo=F}
p1
```

 The tick marks on the xaxis show where each quartile starts. So 25 percent of the countries have 0.2% to 1.4% deaths per infection cases, the next 25 percent have death rates between 1.4% and 2.8%. The third quartile (which includes the United States) have death rates between 2.8% and 4.8%. The final 25 percent have deaths rates between 4.8% and 15.7%.

The countries reporting the lowest death rates:

```{r best rates, echo=F}

survival %>% 
  filter(death_rate<=0.002, death_rate>0) %>% 
  datatable(colnames=c('Country','Total cases','Total deaths','Deaths per cases'),
            options=list(dom='t')) %>% 
  formatCurrency(c('deaths_total','cases_total'),mark=',',digits=0,currency='') %>% 
  formatPercentage('death_rate',digits=1)
  


```

The country reporting the highest death rate:

```{r worst rates, echo=F}

survival %>% 
  filter(death_rate>=0.157, death_rate>0) %>% 
  datatable(colnames=c('Country','Total cases','Total deaths','Deaths per cases'),
            options=list(dom='t')) %>% 
  formatCurrency(c('deaths_total','cases_total'),mark=',',digits=0,currency='') %>% 
  formatPercentage('death_rate',digits=1)
  

```

Some countries that show up frequently in news reports:

```{r countries of interest, echo=F}

coi <- c('China','Germany','Italy','New Zealand','Taiwan','United_States_of_America')

survival %>% 
  filter(countriesAndTerritories %in% coi) %>% 
  datatable(colnames=c('Country','Total cases','Total deaths','Deaths per cases'),
            options=list(dom='t')) %>% 
  formatCurrency(c('deaths_total','cases_total'),mark=',',digits=0,currency='') %>% 
  formatPercentage('death_rate',digits=1)


```

