---
title: Data reporting - more scraping
author: Brian Bowling
date: '2020-06-12'
slug: data-reporting-more-scraping
categories:
  - data_reporting
tags:
  - stats
---

A fellow reporter, Jason Kandel, asked me to look at how to use Google Sheets to scrape wildfire information from this site: https://inciweb.nwcg.gov/accessible-view/

I've had mixed success using Google Sheets as a scraper but I managed to come up with a kludge that requires manual updates as fires get added to the list (which seems to be fairly often based on my limited experience with the site). So this looked like a perfect use of an R script. After some tinkering, here's what I came up with.

## Getting the data

```{r main script, message=F}

# libraries, and what I use them for

# general data wrangling
library(tidyverse)

# scraping
library(rvest)

# date wrangling
library(lubridate)

# data cleaning
library(janitor)

# presenting the data in tables
library(DT)
library(rmarkdown)

# read the main page to get the urls
url_page <- 'https://inciweb.nwcg.gov/accessible-view/'
urls <- read_html(url_page)

# put the info into a tibble
fires_wide <- bind_cols(html_table(urls)[[1]], url =urls %>% html_nodes(xpath='//td[@headers="incident"]') %>% 
  html_nodes('a') %>% 
  html_attr('href')) %>% 
  mutate(url = paste0('https://inciweb.nwcg.gov',url)) %>% 
  clean_names()

# function to scrape each incident page

get_details <- function(fire_name, fire_url) {
  
  print(c(fire_name, fire_url))
  details = tibble()
  
  try({
      incident_page <- read_html(fire_url)
        details <- incident_page %>% 
          html_table() %>% 
          bind_rows() %>%
          rename(category=X1, value=X2) %>% 
          mutate(name = fire_name)
  })
  
  if (nrow(details)==0) {
    details = tibble(category="read error",value="not found",name=fire_name)
  }

  return(details) 
}

# read each of the incident pages and put their info into a single tibble
details_narrow <- bind_rows(map2(fires_wide$incident, fires_wide$url, get_details))

# do a little renaming and rearranging to make it match the fires table terminology

details_narrow <- details_narrow %>% 
  rename(incident = name) %>% 
  select(incident, category, value)


```

The console output started out as a debugging measure because one of the incident pages was down when I was working on the script and it caused an error that halted the script. The printout let me see which url was causing the problem. I kept the output because it gives a nice view of what is now contained in the *fires_wide* tibble.

The error also encouraged me to add some error handling in the form of the *try()* function which allows the script to keep running even if it gets a **404 page not found** response. I didn't put the main link page scrape inside a try statement since there's really no point to running the rest of the script if it can't download the links.

The next step would be to combine some, or possible all, of the info in the *details_narrow* tibble with the *fires_wide* tibble, but first let's talk about wide vs narrow data.

## Wide vs narrow


A wide format is similar to the familiar spreadsheet format where each observation (row) represents a complete record with each piece of information associated with the record in a separate column. It's easy to read but not so easy to analyze.

For example, here's the *fires_wide* tibble:

```{r}

paged_table(fires_wide)

```

A narrow format (also known as "tidy" data in the R community, hence "tidyverse") typically has one observation (row) for each piece of information with the other columns in the observation representing IDs or descriptors that tell you what the information belongs to and what it represents. It's useful for graphing.


For example, the *details_narrow* tibble:

```{r}

paged_table(details_narrow)

```

In addition to being easier to read, the wide format also makes data cleaning easy.

The main issue in the *fires_wide* data is that the size value is character instead of a number so we can't do math on it. I'll extract the number and change it from "size" to "acreage."


```{r}

fires_wide <- fires_wide %>% 
  mutate(size = as.numeric(str_extract(size, '\\d+'))) %>% 
  rename(acreage=size)

paged_table(fires_wide)

```


The *details_narrow* data has some problems, so I'm going to convert it to a wide format and cleanup the resulting column names to make them easier to work with.


```{r}

details_wide <- details_narrow %>% 
  pivot_wider(id_cols=incident, names_from=category, values_from=value) %>% 
  clean_names()


paged_table(details_wide)
```


Just from eyeballing the information, I can see a few problems. One is that the latitude and longitude of each fire is combined into a single record "coordinates." To use the data for mapping, those need to be pulled into their own categories.

Another is that the "current_as_of" values are a javascript statement designed to convert a date string to a date value. It would be more useful to extract the date value, get rid of the rest, and then convert the date value into a date variable. I'll skip that last step, however, since it would prevent me from converting the data back to a narrow format.

### Latitude and longitude

Taking a look at the values in the coordinates category, there is an obvious pattern: <number> latitude, -<number> longitude. We can use str_extract() to pull out the numbers.



```{r}

details_wide <- details_wide %>% 
  mutate(latitude = str_extract(coordinates,pattern='\\d+\\.\\d+'),
         longitude = str_extract(coordinates,pattern='-\\d+\\.\\d+'))

```

The current_as_of values also have a clear pattern. In particular, the info we want is between two double-quote marks. A further complication is the format for the time which includes the -05:00 to show the offset from UTC.

```{r}

details_wide <- details_wide %>% 
  mutate(current_as_of = str_extract(current_as_of,pattern='\\".*?\\"'))

```



I don't see any other data issues other than a lot of missing values, so I'm going to combine the the two wide tibbles. If I need narrow info, such as for graphing, I'll extract it out.


```{r}

fire_info_wide <- merge(fires_wide, details_wide,by='incident')

```


```{r}
paged_table(fire_info_wide)

```

Using the data to answer some questions:

How many acres are on fire in each state?

```{r, message=F}

fire_info_wide %>% 
  group_by(state) %>% 
  summarise(total_acreage = sum(acreage)) %>% 
  datatable(options=list(dom='t'))

```

How many acres by fire type in each state?

```{r, message=F}

fire_info_wide %>% 
  group_by(state, type) %>% 
  summarise(total_acreage = sum(acreage)) %>% 
  datatable(options=list(dom='t'))

```

There's some more cleanup that could be done, such as data entered as characters that could be converted to dates or numbers, but this enough for now.

